---
title: "Final paper"
author: "Alyssa Hagele, Abby Malech, Henderson Vo"
format: pdf 
execute:
  message: false
  warning: false
  echo: false
bibliography: references.bib
---

```{r}
#| label: load-packages

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(naniar)
library(patchwork)
library(gghighlight)
```

```{r}
#| label: load-data
# load vegetation data
veg <- read_csv(here("data", "veg.csv"))
# load vegetation metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

# Introduction

Restoration efforts at the North Campus Open Space (NCOS) are trying to increase species richness across major habitats, where key species are driving changes in composition, habitat stability, and biodiversity. 10 native and non-native plants were selected for this study based on high observations, presence during all monitoring years (2021-2025), and frequent mention in the 2024 monitoring report and native plant catalog. This study specifically looks at how species responded to a prescribed burn that happened in Fall 2023 as well as the effects of a year with heavier rainfall in 2024. Therefore, the main questions for this study include, **“how has percent cover of key native species and overall native species richness changed through time in grasslands compared to non-native species?”** and **“What impact does the 2023 fire and 2024 water year have on species richness and cover for both native and non-native grassland species?”**

Our native species of interest are:

`Deinandra fasciculata`,\ `Distichlis spicata`,\ `Erigeron canadensis`,\ `Stipa pulchra`, `Symphyotrichum subulatum`,

And our non-native species of interest are:

`Bromus hordeaceus`,\ `Festuca myuros`,\ `Festuca perennis`,\ `Medicago polymorpha`,\ `Polypogon monspeliensis`.

This study will pay close attention to Stipa pulchra — California's state grass — as well as Festuca perennis, a non-native grass, as these species are greatly abundant across all eight grassland transects. Lastly, Deinandra fasciculata is a species endemic to the Santa Barbara region and was stated to have “responded well to the fire”, which might indicate that native species perform better after a prescribed burn [@rickard2024]. The monitoring report also stated that S. pulchra was predicted to respond well to the fires. Other factors besides native/non-native status will be taken into account in this study, such as perennial versus annual grasses. Monitoring plant cover and richness over time can best inform restoration experts how certain intentional or unintentional events affect overall restoration progress. A study conducted on a grassland habitat found that species show positive responses to water availability extremes as net primary productivity (NPP) increased during wet years. Additionally, functional diversity was considered an indicator of healthier ecosystems and more stable NPP, which can be informed by *structural* diversity, such as total cover and richness [@alexander2023]. This is significant in the case of NCOS, as increasing native species abundance and diversity can enhance ecosystem health and resilience.

# Methods

**Data Collection and Study Area**

To answer our research question, we used a vegetation dataset sourced from the Cheadle Center for Biodiversity and Ecological Restoration (CCBER), and provided to us by the ENVS 193DD course. This dataset spans 5 years (2021-2025) of vegetation data from across several habitats in North Campus Open Space (NCOS) including grasslands, mosaic habitats, wetlands, open water, woodlands, and shrublands. We chose to focus on native perennial grasslands, which encompasses measurements from 8 different transects across the 16.8-acre mesa. The method of data collection used for vegetation surveys is the Quadrat Transects (QT) method. This method involves the use of 11 one-square-meter quadrats placed by a 30-meter transect, with quadrats alternating between the left and right side of the transect line every 3 meters. “The quadrats are subdivided into 100 ten-centimeter squares and the percent cover is estimated \[by human monitors\] for each species in the quadrat,” (NCOS Restoration Project Monitoring Report).

**Analytical Approach**

We will not be performing a formal statistical analysis for our project because we are primarily exploring a correlative relationship rather than predicting trends for species cover and richness. However, we will be identifying major events that might influence a directional response in vegetation percent cover and species richness, and examining whether a response was captured in our data set. Major events include the prescribed burn in Fall 2023 because of its effects as an intermediate disturbance event, as well as wet (2023, 2024) and dry (2021, 2022) water years because water is a limiting resource for key native grass species (S. pulchra). These factors are potential influences on how percent cover and species richness change over time. This informal analysis will be conducted by examining the visual components of our figures. These visual components include changes in size or direction of geometries representing key variables, as well as geometries marking important contextual events.

# Results

```{r}
#| label: implicit-data
# create a new object from veg
gl_implicitNA <- veg |> 
  # filter all occurrences where transect_name contains GL
  filter(str_detect(transect_name, "GL") &
           site == "ncos") |> 
  # group by year, pool, and species
  group_by(year, transect_name, psoc) |>
  # calculate mean percent cover
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join with metadata by year_pool
  left_join(metadata, by = "year_pool")
```

```{r}
#| label: explicit-data

# creating new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains GL
  filter(str_detect(transect_name, "GL") &
           site == "ncos") |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)
```

```{r}
#| label: alyssa wrangling for vis 1
# create a new object from veg
gl_implicitNA_alyssa <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL") &
           site == "ncos") |> 
  # group by year, pool, and species
  group_by(year, transect_name, psoc, cover_category) |>
  # calculate mean percent cover
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join with metadata by year_pool
  left_join(metadata, by = "year_pool")

# create a new object
grass_cover <- gl_implicitNA_alyssa |>
  # filter to only include native and non-native cover grasses
  filter(cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))

# create a new object
gl_transects <- gl_implicitNA_alyssa |>
  filter( # filter to only include grassland transects
    transect_name %in% c("GL-02", "GL-07", "GL-09", "GL-10",
                         "GL-11", "GL-12", "GL-14", "GL-15"),
    # filter to only include native and non-native cover observations
    cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

```{r}
#| label: native-vector

# creating new vector from veg
native_spp <- veg |> 
  # filter all occurrences where transect_name contains GL
  filter(str_detect(transect_name, "GL") &
           site == "ncos" &
           cover_category == "NATIVE COVER") |> 
# pull to extract data frame column as a vector
  pull(psoc)
```

```{r}
#| label: invasive-vector

# creating new vector from veg
invasive_spp <- veg |> 
  # filter all occurrences where transect_name contains GL and cover category is non-native
  filter(str_detect(transect_name, "GL") &
           site == "ncos" &
           cover_category == "NON-NATIVE COVER") |> 
# pull to extract data frame column as a vector
  pull(psoc)
```

```{r}
#| label: native and non native species list


# native species
# species most significant: Stipa_pulchra and Distichlis_spicata
native_interest <- c(
  "Stipa_pulchra",
  "Distichlis_spicata",
  "Deinandra_fasciculata",
  "Erigeron_canadensis",
  "Symphyotrichum_subulatum"
)

# invasive / non-native species
# species most significant: Festuca_perennis and Polypogon_monspeliensis
non_native_interest <- c(
  "Festuca_perennis",
  "Polypogon_monspeliensis",
  "Bromus_hordeaceus",
  "Festuca_myuros",
  "Medicago_polymorpha"
)

# all selected species
all_interest <- c(
  "Stipa_pulchra",
  "Distichlis_spicata",
  "Deinandra_fasciculata",
  "Erigeron_canadensis",
  "Symphyotrichum_subulatum",
  "Festuca_perennis",
  "Polypogon_monspeliensis",
  "Bromus_hordeaceus",
  "Festuca_myuros",
  "Medicago_polymorpha"
)
```

```{r}
#| label: alyssa wrangling for vis 2 

# create a new object from species lists
selected_species <- c(native_interest, non_native_interest)

# create a new object from grasslands data
selected_species_cover <- gl_implicitNA |>
  # filter to only include species of interest
  filter(psoc %in% selected_species) |>
  # classify each species as native or non native
  mutate(
    species_group = case_when(
      psoc %in% native_interest ~ "Native",
      psoc %in% non_native_interest ~ "Non-native"))
```

```{r}
#| label: key four spp wrangling

# choose four focal species
four_species <- c("Deinandra_fasciculata",
                  "Stipa_pulchra",
                  "Festuca_perennis",
                  "Polypogon_monspeliensis")

# create object with only the four focal species
four_species_cover <- selected_species_cover |>
  # filter to only include four focal species
  filter(psoc %in% four_species) |>
  # group by year, species, and native classification
  group_by(year, psoc, species_group) |>
  # calculate total cover
  summarize(total_cover = sum(mean_pc, na.rm = TRUE),
            .groups = "drop") |>
  # ungroup
  ungroup()
```

```{r}
#| label: native-richness

# create new object from gl_explicit0
native_species_rich <- gl_explicit0 |> 
# group by year and pool
  group_by(year, transect_name) |> 
# filter data frame with vector into get native species, 
# also filter percent cover to identify species presence
  filter(psoc %in% native_spp, mean_pc > 0) |> 
# species richness calculation
  summarize(spp_rich = n_distinct(psoc)) |> 
  # ungroup data frame
  ungroup()
```

```{r}
#| label: non-native-richness

# create new object from gl_explicit0
invasive_species_rich <- gl_explicit0 |> 
# group by year and pool
  group_by(year, transect_name) |> 
# filter data frame with vector into get native species, 
# also filter percent cover to identify species presence
  filter(psoc %in% invasive_spp, mean_pc > 0) |> 
# species richness calculation
  summarize(spp_rich = n_distinct(psoc)) |> 
  # ungroup data frame
  ungroup()
```

```{r}
#| label: alyssa vis-3 species richness native non native wrangling

# combine native and non-native richness data
all_species_richness <- bind_rows(
  # add native label
  native_species_rich |>
    mutate(type = "Native"),
  # add non-native label
  invasive_species_rich |>
    mutate(type = "Non-native")
)
```

```{r}
#| label: figure-1
#| fig-height: 5
#| fig-width: 7

# combine native and non-native richness data
all_species_richness <- bind_rows(
  native_species_rich |>
  # create native column
    mutate(type = "Native"),
  invasive_species_rich |>
  # create non-native column
    mutate(type = "Non-native")
)
# summarize species richness by year and type
species_richness_summary <- all_species_richness |>
  # group by year, stacked bar by type
  group_by(year, type) |>
  # calculate total richness column
  summarize(
    total_richness = sum(spp_rich, na.rm = TRUE),
    .groups = "drop"
  )
# create stacked bar plot over time
species_richness_bar_plot <- species_richness_summary |>
  # adding ggplot base layer
  ggplot(aes
      # x-axis: year
      # y-axis: total richness
      # fill by native/non-native
         (x = year,
             y = total_richness,
             fill = type)) +
  # stacked bars
  geom_col() +
  # remove space between x-axis
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 210)) +
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray50",
    linewidth = 1
  ) +
  # label prescribed fire event
  annotate(
    "text",
    x = 2023.5,
    y = 184,
    label = "Prescribed Fire\n(Fall 2023)",
    angle = 90,
    vjust = -0.5,
    size = 3,
    color = "gray40"
  ) +
  # adding custom colors
  scale_fill_manual(values = c(
    "Native" = "#4A9782",
    "Non-native" = "#D16D5B"
  )) +
  # adding custom labels
  labs(
    title = "Native vs Non-native Species Richness Over Time",
    x = "Year",
    y = "Species Richness",
    fill = "Species Type"
  ) +
  # changing from default theme
  theme_bw() +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

# display plot
species_richness_bar_plot
```

**Figure 1: Native species richness maintained similar levels while non-native species richness declined over time.** The x-axis represents time in years. The y-axis measures species richness across all transects for all species (not just selected species). Each column represents the total species count in a particular year. The dashed line marks the relative point in time between surveys when the prescribed burn occurred. 

```{r}
#| label: figure-2
#| fig-width: 10
#| fig-height: 6

# create a new object 
selected_pc_native <- selected_species_cover |>
  # group by year, species, and species group
  group_by(year, psoc, species_group) |>
  # filter to only include native species
  filter(species_group == "Native") |> 
  # calculate total cover
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  # ungroup
  ungroup() |>
  # replace underscores with spaces in species names
  mutate(psoc = str_replace_all(psoc, "_", " ")) |>
  # arrange plot in order of descending total cover
  arrange(desc(total_cover)) |>
  # convert species data type into factor
  mutate(psoc = factor(psoc, levels = unique(psoc))) |>
  # base layer: ggplot, 
  # set x- and y-axis and differentiate by species group
  ggplot(aes(x = year,
             y = total_cover,
             color = species_group)) +
  # first layer: line geometry
  geom_line(aes(group = psoc),
            linewidth = 1.2) +
  #second layer: highlight native species line
  gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(
      color = "gray70",
      alpha = 0.6,
      linewidth = 0.5
    )
  ) +
  #custom native spp color 
  scale_color_manual(
    breaks = c("Native"),
    values = c(
      "Native" = "#4A9782"
    )
  ) +
  #seperate panels per species
  facet_wrap(~ psoc, scales = "free_x", nrow = 1) +
  #set x axis breaks + spacing
  scale_x_continuous(
    breaks = 2021:2025,
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  #relabel title, x and y axis, legend 
  labs(
    title = "Selected Native Species Percent Cover Over Time",
    x = "Year",
    y = "Total Percent Cover",
    color = "Species Group"
  ) +
  #custom theme 
  theme_bw() + 
  #change x axis text size
  theme(axis.text.x = element_text(size = 8)) +
  theme(strip.text.x = element_text(size = 7.5)) +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) +
  #remove legend
  theme(legend.position = "none")

#create new object from selected_species_cover
selected_pc_nonnative <- selected_species_cover |>
  #group by year, species, species group 
  group_by(year, psoc, species_group) |>
  #filter to only include invasives
  filter(species_group == "Non-native") |>
  #calculate total percent cover for each species
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  #ungroup data frame
  ungroup() |>
  #replace dashes in species names with spaces
  mutate(psoc = str_replace_all(psoc, "_", " ")) |> 
  #arrange species by descending total cover
  arrange(desc(total_cover)) |>
  #convert species names into factor to preserve order
  mutate(psoc = factor(psoc, levels = unique(psoc))) |>
  #base layer: ggplot 
  #x axis is year, y axis is total_cover, color by species group 
  ggplot(aes(x = year,
             y = total_cover,
             color = species_group)) +
  #first layer: lines
  geom_line(aes(group = psoc),
            linewidth = 1.2) +
  #second layer: highlight invasive spp lines
  gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(
      color = "gray70",
      alpha = 0.6,
      linewidth = 0.5
    )
  ) +
  #custom color for invasive species 
  scale_color_manual(
    breaks = c("Non-native"),
    values = c(
      "Non-native" = "#C96B5C"
    )
  ) +
  #separate panels for each species 
  facet_wrap(~ psoc, scales = "free_x", nrow = 1) +
  scale_x_continuous(
    breaks = 2021:2025,
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  #relabeling title, x and y axis
  labs(
    title = "Selected Non-native Species Percent Cover Over Time",
    x = "Year",
    y = "Total Percent Cover",
    color = "Species Group"
  ) +
  #custom theme
  theme_bw() + 
  #change x axis text size
  theme(axis.text.x = element_text(size = 8)) + 
  theme(strip.text.x = element_text(size = 7.5)) +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) +
  #remove legend 
  theme(legend.position = "none")

#combine native and non-native plots into one figure
selected_pc_native/selected_pc_nonnative
```

**Figure 2: On average, native vegetation appears to decrease or maintain their level of percent cover over time while non-native vegetation increase or maintain their level of percent cover over time.** The x-axis represents time in years. The y-axis measures total percent cover of vegetation across all quadrats in all transects. Vertices represent individual measurements. Line segments represent the change in total percent cover between years. Each panel shows the change in total percent cover over time for an individual species.

```{r}
#| label: figure-3
#| fig-width: 10
#| fig-height: 7

gl_transects |>
  # group data frame by year, transect, and cover category
  group_by(year, transect_name, cover_category) |>
  # calculate total cover
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  # base layer: ggplot
  # x axis is year, y is total_cover, color by cover category
  ggplot(aes(x = year,
             y = total_cover,
             color = cover_category)) +
  # first layer: lines
  geom_line() +
  # second layer: points
  geom_point() +
  #panels by transect, set x scale to free
  facet_wrap(~ transect_name, scales = "free_x") +
  # relabel title, x and y axis, legend
  labs(
    title = "Native vs Non-Native Percent Cover Over Time Across GL Transects",
    x = "Year",
    y = "Total Percent Cover",
    color = "Cover Category"
  ) +
  # custom color
  scale_color_manual(
    values = c("NON-NATIVE COVER" = "#C96B5C",
               "NATIVE COVER" = "#4A9782")) +
  # change theme
  theme_bw() +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

**Figure 3: All transects besides 9 and 14 decrease or maintain their level of native percent cover, while non-native percent cover increases over time in all transects. Following the 2023 prescribed burn, percent cover from 2023 to 2024 increased only for transects 14 and 15 while decreasing or maintaining similar levels in the other 6 transects.** The x-axis represents time in years. The y-axis measures mean percent cover for each individual transect, grouped by all native vs non-native species (not just selected species).

```{r}
#| label: stipa-jitter-plot

# new object from selected_species_cover
stipa_plot <- selected_species_cover |>
  #filter for only Stipa pulchra
  filter(psoc == "Stipa_pulchra") |>
  #base layer: ggplot
  #x axis is year, y is mean_pc
  ggplot(aes(x = year,
             y = mean_pc)) +
  #layer 1: blue shaded area showing 2024 water year
  #customize shaded area
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
  #second layer: transect observations / jitter plot
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
  #third layer: mean observations by year
stat_summary(
  aes(color = "Native"),
  fill = "#4A9782", 
  geom = "point",
  fun = "mean",
  shape = 21,
  size = 5,
  stroke = 1.2
) +
  #customize species type color 
  scale_color_manual(
  name = "Species Type",
  values = c(
    "Native" = "#4A9782",
    "Non-native" = "#9B5DE5"
  )
) +
  #fourth layer: line showing prescribed fire
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
  #fifth layer: prescribed fire label 
  annotate(
    "text",
    x = 2023.55,
    y = 50,
    label = "Prescribed fire\n(Fall 2023)",
    hjust = 0,
    vjust = 1,
    color = "gray35",
    size = 3.5
  ) +
  #show every year on x-axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
  #relabel title, y axis, no x axis title
  labs(
    title = "Stipa Pulchra",
    y = "Mean Percent Cover",
    x = NULL
  ) +
  # custom theme
  theme_bw() +
  #customize plot 
  theme(
    plot.title = element_text(face = "bold",
                              size = 16),
    plot.subtitle = element_text(color = "gray35"),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: deinandra-jitter-plot

#new object from selected_species_cover
deinandra_plot <- selected_species_cover |>
  #keep only Deinandra fasciculata
  filter(psoc == "Deinandra_fasciculata") |>
  #base layer: ggplot
  ggplot(aes(x = year,
             y = mean_pc)) +
#first layer: shaded area showing 2024 water year
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
#second layer: points showing individual observations
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
#third layer: mean yearly observations
  stat_summary(
    aes(color = "Native"),
    fill = "#4A9782",
    geom = "point",
    fun = "mean",
    shape = 21,
    size = 5,
    stroke = 1.2
  ) +
#custom colors per species type
  scale_color_manual(
    name = "Species Type",
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#9B5DE5"
    )
  ) +
  #fourth layer: line showing prescribed fire
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
#show every year on x axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
#relabel title, no x and y axis label 
  labs(
    title = "Deinandra Fasciculata",
    y = NULL,
    x = NULL
  ) +
  # custom theme
  theme_bw() +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: festuca-jitter-plot

#new object from selected_species_cover
festuca_plot <- selected_species_cover |>
  #filter to keep only Festuca perennis
  filter(psoc == "Festuca_perennis") |>
  #base layer: ggplot
  #x axis is year, y is mean_pc
  ggplot(aes(x = year,
             y = mean_pc)) +
  #first layer: blue shaded area represents water year 2024
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
#second layer: individual points for each observation
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
#third layer: mean points per year
  stat_summary(
    aes(color = "Non-native"),
    fill = "#9B5DE5",
    geom = "point",
    fun = "mean",
    shape = 21,
    size = 5,
    stroke = 1.2
  ) +
#custom colors per species type
  scale_color_manual(
    name = "Species Type",
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#9B5DE5"
    )
  ) +
  #fourth layer: line for prescribed fire event
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
#show every year on x axis 
  scale_x_continuous(
    breaks = 2021:2025
  ) +
#relabel title, add x and y axis label 
  labs(
    title = "Festuca Perennis",
    y = "Mean Percent Cover",
    x = "Year"
  ) +
  # custom theme
  theme_bw() +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: polypogon-jitter-plot

#new object from selected_species_cover
polypogon_plot <- selected_species_cover |>
  #filter to keep only Polypogon monspeliensis
  filter(psoc == "Polypogon_monspeliensis") |>
#base layer: ggplot
#x axis is year, y is mean_pc
  ggplot(aes(x = year,
             y = mean_pc)) +
  #first layer: blue shaded area shows 2024 water year
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
#second layer: points per each individual observation
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
#third layer: mean points per year
  stat_summary(
    aes(color = "Non-native"),
    fill = "#9B5DE5",
    geom = "point",
    fun = "mean",
    shape = 21,
    size = 5,
    stroke = 1.2
  ) +
#custom colors for species type 
  scale_color_manual(
    name = "Species Type",
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#9B5DE5"
    )
  ) +
  #fourth layer: line for prescribed fire
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
  #show every year on x-axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
#relabel title, no y axis label, adding x axis label 
  labs(
    title = "Polypogon Monspeliensis",
    y = NULL,
    x = "Year"
  ) +
  # custom theme
  theme_bw() +
  #customize plot
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: figure-4
#| fig-width: 10
#| fig-height: 8

#creating new plot combined last four plots
combined_species_plot <-
  ((stipa_plot + deinandra_plot) /
   (festuca_plot + polypogon_plot)) +
  #collect legends into one shared legend
  plot_layout(guides = "collect") +
  #add overall title and subtitle
  #customizing plot / title and subtitle text size and color
  plot_annotation(
    title = "Percent Cover of Four Keys Species Through Time",
    subtitle = "Blue shading indicates the 2024 water year (143% of normal precipitation).",
    theme = theme(
      plot.title = element_text(
        size = 20,
        face = "bold",
        hjust = 0.5
      ),
      plot.subtitle = element_text(
        size = 12,
        color = "gray35",
        hjust = 0.5
      )
    )
  ) &
  #move shared legend to bottom
  theme(
    legend.position = "bottom"
  )

# display plot
combined_species_plot
```

**Figure 4: Mean percent cover increased for species with low abundance and decreased for species with high abundance prior to the fire event. This was not limited to native versus non-native status. For example, S. pulchra (native) and F. Perennis (non-native) mean percent cover both decreased after the burn.** The x-axis represents time in years. The y-axis measures mean percent cover across all transects. Grey circles represent individual observations, while the green and purple dots mark the mean of those individual points. The dashed line marks the relative point in time between surveys when the prescribed burn occurred, and the blue highlighted section is a wetter-than-average year (2024).

**Percent Cover and Species Richness Trends**

Overall, our results show that both native and non-native species richness increased following the prescribed burn of fall 2023 and during the 2024 water year, with non-native richness having the largest increase. Concerning percent cover, we saw native percent cover increase in transects 14 and 15 after the burn, which are the closest transects to vernal pools 1 and 2. Non-native percent cover after the prescribed burn increased in transects 7, 10, 11, and 15, with non-native cover generally remaining relatively high across most transects during the study period. Spatially, as mentioned, we found that native species percent cover increased most strongly near vernal pools 1 and 2, which is significant because these vernal pools also experienced the greatest water depths during the 2024 water year. Lastly, we found that non-native percent cover and overall species richness increased most strongly in the southeastern grassland transects.

**Focal Species Trends**

We also found several meaningful patterns concerning our focal species. The percent cover of Stipa pulchra, a key NCOS restoration target species, increased from 2021 to 2023; however, it declined after the burn and the 2024 water year, overall decreasing over time. The percent cover of Deinandra fasciculata, a species endemic to Santa Barbara, remained relatively low throughout the study period; however, percent cover increased slightly following the prescribed burn and during the 2024 water year. Concerning non-native focal species, we found that Festuca perennis, another restoration target, decreased slightly after the burn; however, it rebounded in 2025 and now has one of the highest percent covers out of any species. Polypogon monspeliensis did not seem to react to the burn as it increased slightly in 2024; however, it may have been out-competed by other species during the 2024 water year, as it declined substantially in 2025 and now has a very small percent cover.

# Discussion

**Connection to Existing Publications**

Concerning Stipa pulchra, our results supported other relevant studies that also determined that prescribed fires had little positive effect on Stipa pulchra, a key NCOS restoration target [@keeley2023]. Its substantial decline in percent cover after the prescribed burn of fall 2023 is likely due to being outcompeted by invasive opportunistic annuals who are establishing faster following disturbances [@gornish2017; @hamilton1999]. As said in the monitoring report, it will likely take longer observation to determine whether declines in Stipa pulchra represent short-term responses to disturbance events or long-term trends. While we did see Festuca perennis decline slightly after the fire, it did rebound in 2025 and has increased over time; however, the rate of its increase has slowed after multiple disturbance events. Other grassland studies also found that Festuca perennis did increase after burns; however, it did have negative trends following large precipitation events [@keeley2023]. This could possibly be the reason why its rate of increase in percent cover has slowed after 2023 (because of the heavy 2024 water year). Our results also concur with the Gornish & Shaw study, which concluded that Deinandra fasciculata responds well to disturbances, as we have seen an overall increase in its relatively low percent cover over time, especially after the prescribed burn.

Our results concerning how native and non-native species reacted to disturbance events such as the fall 2023 prescribed burn and 2024 water year followed similarly to existing studies. Among our key species, we consistently found that non-native annuals were outcompeting native perennials, as trends between the two were often inverse [@corbin2010; @gornish2017]. This is due to annual species with shallow root systems often being able to establish quicker after disturbance events. Even among our key perennial and annual species, without distinguishing between native and non-native species, we found generally annual species were outperforming perennials; for example, even though we saw Stipa pulchra decline, Deinandra fasciculata, an annual native, has increased over time. One notable exception, however, is Distichlis spicata, which is a native perennial that has been thriving at NCOS. This is likely due to it being a very salt-, drought-, and stress-tolerant species that has adapted to many California habitats [@pessarakli2012].

**Relevance to Other Ecosystems**

Our results allow us to make inferences about California grassland ecosystems. The patterns exhibited as NCOS have been well-documented among several California grassland studies; disturbance events such as fires, sheep grazing, and wet water years often favor fast-growing annuals over slow-establishing perennial grasses. The decline of Stipa pulchra and the increase of fast-growing annuals reflect the overall pattern of California grasslands in which annuals are outcompeting native perennials. One new pattern we found from our study is that native percent cover generally increased in transects located near vernal pools 1 and 2, which are the deepest vernal pools at NCOS that had the highest water input during the 2024 water year. This suggests that greater water availability may be benefiting native species establishment and that ecosystems with deeper vernal pools may be more beneficial to native perennial species because deeper root systems may have greater access to water.

**Limitations**

There are certain limitations to consider that could affect the results of our study. First, the vegetation data is a survey that is taken annually during July, rather than multiple times throughout the year. Therefore, seasonal fluctuations were not able to be visualized in the analysis. The nature of restoration monitoring involves subjectivity during observations when cover is estimated visually from each transect. This means that measured cover can be easily over- or underestimated. Additionally, the prescribed fire only occurred once during the time that the vegetation data set was recorded. While we did see evidence of the fire having a more equitable effect on native and non-native species alike, more events in the future should be taken into consideration to reach an interpretation with higher confidence. It is hard to establish causation considering that the fire occurred in the Fall of 2023 and vegetation measurements were not recorded until July of 2024. Other factors like annuals versus perennials, proximity to water abundance (vernal pools), and proximity to trails with increased human disturbances were not captured explicitly within the vegetation data set but are acknowledged in this study.

# References
