---
title: "in-class-code"
format: pdf
execute: 
  message: false
  warning: false
---

# Packages

```{r}
#| label: load-packages

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(naniar)
library(patchwork)
library(gghighlight)
```

# Data

```{r}
#| label: load-data
veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

# Class code

```{r}
#| label: implicit-data
# create a new object from veg
gl_implicitNA <- veg |> 
  # filter all occurrences where transect_name contains GL
  filter(str_detect(transect_name, "GL") &
           site == "ncos") |> 
  # group by year, pool, and species
  group_by(year, transect_name, psoc) |>
  # calculate mean percent cover
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join with metadata by year_pool
  left_join(metadata, by = "year_pool")
```

```{r}
#| label: explicit-data

# creating new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains GL
  filter(str_detect(transect_name, "GL") &
           site == "ncos") |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)
```

```{r}
#| label: native-species-explicit%cover

# creating new object from veg
gl_native_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains GL & native
  filter(str_detect(transect_name, "GL") &
           site == "ncos" &
           cover_category == "NATIVE COVER") |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)
```

```{r}
#| label: invasive-species-explicit%cover

# creating new object from veg
gl_invasive_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains GL & native
  filter(str_detect(transect_name, "GL") &
           site == "ncos" &
           cover_category == "NON-NATIVE COVER") |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)
```

```{r}
#| label: missingness-plots

gg_miss_var(gl_implicitNA)

gg_miss_var(gl_explicit0)
```

No missing values in either data frame.

# Grasslands vectors

```{r}
#| label: native-vector

# creating new vector from veg
native_spp <- veg |> 
  # filter all occurrences where transect_name contains GL
  filter(str_detect(transect_name, "GL") &
           site == "ncos" &
           cover_category == "NATIVE COVER") |> 
# pull to extract data frame column as a vector
  pull(psoc)

# sample of ten rows
sample(native_spp, 10)
```

```{r}
#| label: invasive-vector

# creating new vector from veg
invasive_spp <- veg |> 
  # filter all occurrences where transect_name contains GL and cover category is non-native
  filter(str_detect(transect_name, "GL") &
           site == "ncos" &
           cover_category == "NON-NATIVE COVER") |> 
# pull to extract data frame column as a vector
  pull(psoc)

# sample of ten rows
sample(invasive_spp, 10)
```

# Species richness

```{r}
#| label: native-richness

# create new object from gl_explicit0
native_species_rich <- gl_explicit0 |> 
# group by year and pool
  group_by(year, transect_name) |> 
# filter data frame with vector into get native species, 
# also filter percent cover to identify species presence
  filter(psoc %in% native_spp, mean_pc > 0) |> 
# species richness calculation
  summarize(spp_rich = n_distinct(psoc)) |> 
  # ungroup data frame
  ungroup()

# sample 10 rows
slice_sample(native_species_rich, n = 10)
```

```{r}
#| label: non-native-richness

# create new object from gl_explicit0
invasive_species_rich <- gl_explicit0 |> 
# group by year and pool
  group_by(year, transect_name) |> 
# filter data frame with vector into get native species, 
# also filter percent cover to identify species presence
  filter(psoc %in% invasive_spp, mean_pc > 0) |> 
# species richness calculation
  summarize(spp_rich = n_distinct(psoc)) |> 
  # ungroup data frame
  ungroup()

# sample 10 rows
slice_sample(invasive_species_rich, n = 10)
```

# Outliers???

```{r}

```

# Total Cover

```{r}
#| label: total-cover-native-veg

# create new object from veg
native_pc <- veg |>
  # filter only native species
  filter(psoc %in% native_spp) |>
  # filter only grasslands at NCOS
  filter(str_detect(transect_name, "GL") & site == "ncos") |>
  # group by year and transect
  group_by(year, transect_name) |> 
  # calculate implicit missingness sum
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # united column to match the one in metadata
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join metadata
  left_join(metadata, by = "year_pool") |> 
  # dividing total % cover by the # of quadrats in the observed year
  mutate(total_pc = sum_pc/num_quad)
  
  # sample of 10
slice_sample(native_pc, n = 10)
```

```{r}
#| label: total-cover-non-native-veg

# create new object from veg
invasive_pc <- veg |>
  # filter only invasive species
  filter(psoc %in% invasive_spp) |>
  # filter only grasslands at NCOS
  filter(str_detect(transect_name, "GL") & site == "ncos") |>
  # group by year and transect
  group_by(year, transect_name) |> 
  # calculate implicit missingness sum
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # united column to match the one in metadata
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join metadata
  left_join(metadata, by = "year_pool") |> 
  # dividing total % cover by the # of quadrats in the observed year
  mutate(total_pc = sum_pc/num_quad)
  
  # sample of 10
slice_sample(invasive_pc, n = 10)
```


# Interest species wrangling

```{r}
#| label: native and non native species list


# native species
# species most significant: Stipa_pulchra and Distichlis_spicata
native_interest <- c(
  "Stipa_pulchra",
  "Distichlis_spicata",
  "Deinandra_fasciculata",
  "Erigeron_canadensis",
  "Symphyotrichum_subulatum"
)

# invasive / non-native species
# species most significant: Festuca_perennis and Polypogon_monspeliensis
non_native_interest <- c(
  "Festuca_perennis",
  "Polypogon_monspeliensis",
  "Bromus_hordeaceus",
  "Festuca_myuros",
  "Medicago_polymorpha"
)

# all selected species
all_interest <- c(
  "Stipa_pulchra",
  "Distichlis_spicata",
  "Deinandra_fasciculata",
  "Erigeron_canadensis",
  "Symphyotrichum_subulatum",
  "Festuca_perennis",
  "Polypogon_monspeliensis",
  "Bromus_hordeaceus",
  "Festuca_myuros",
  "Medicago_polymorpha"
)
```

```{r}
#| label: filtering-interest-species

gl_native_interest <- gl_native_explicit0 |> 
  filter(psoc %in% native_interest) |> 
  mutate(native = "native")

gl_invasive_interest <- gl_invasive_explicit0 |> 
  filter(psoc %in% non_native_interest) |> 
  mutate(native = "non_native")

gl_all_interest <- bind_rows(gl_native_interest, gl_invasive_interest)
```


```{r}
#| label: wrangling-vis-1
# data frame for vis 1
gl_all_pc <- gl_all_interest  |> 
  group_by(psoc, year, native) |> 
  summarize(sum_pc = sum(sum_pc),
            sum_quad = sum(num_quad)) |> 
  ungroup() |> 
  mutate(mean_pc = sum_pc/sum_quad)
```

```{r}
#| label: wrangling-vis-2
# tbd

```


```{r}
#| label: wrangling-vis-3
# data frames for vis 3
gl_native_tc <- gl_native_interest |> 
  group_by(psoc, year_pool, native) |> 
  summarize(sum_pc = sum(sum_pc)) |> 
  ungroup()

gl_invasive_tc <- gl_invasive_interest |> 
  group_by(psoc, year_pool, native) |> 
  summarize(sum_pc = sum(sum_pc)) |> 
  ungroup()
```

```{r}
#| label: wrangling-vis-4
gl_stacked_native <- gl_all_interest |> 
  group_by(native, year) |> 
  summarize(mean_pc_sum = sum(mean_pc)) |> 
  ungroup() |> 
  mutate(native)

```


# Visualizations

```{r}
#| label: vis-1
vis1 <- ggplot(data = gl_all_pc,
       aes(x = year,
           y = mean_pc,
           color = psoc)) +
  geom_line() +
  facet_wrap(~ native)

vis1
```

```{r}
#| label: vis-3

native_tc_plot <- ggplot(data = gl_native_tc,
       aes(x = psoc,
           y = sum_pc,
           color = psoc)) +
  geom_jitter(height = 0,
              width = 0.2) +
  stat_summary(geom = "pointrange",
               # argument to calculate standard error
               fun.data = mean_sdl,
               fun.args = list(mult = 1),
               size = 1,
               linewidth = 1,
               color = "red") +  
  labs(title = "Native Total Cover") +
  theme(legend.position = "none")

invasive_tc_plot <- ggplot(data = gl_invasive_tc,
       aes(x = psoc,
           y = sum_pc,
           color = psoc)) +
  geom_jitter(height = 0,
              width = 0.2) +
  stat_summary(geom = "pointrange",
               # argument to calculate standard error
               fun.data = mean_sdl,
               fun.args = list(mult = 1),
               size = 1,
               linewidth = 1,
               color = "red") +  
  labs(title = "Non-native Total Cover") +
  theme(legend.position = "none")

vis3 <- native_tc_plot + invasive_tc_plot
vis3
```

```{r}
#| label: vis-4

vis4 <- ggplot(data = gl_stacked_native,
               aes(x = year,
                   y = mean_pc_sum,
                   fill = native)) +
  geom_area()

vis4
```

## alyssa visualizations

### non native and native pc over time FACETED by transect
```{r}
#| label: alyssa wrangling for vis 1

gl_implicitNA_alyssa <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL") &
           site == "ncos") |> 
  # group by year, pool, and species
  group_by(year, transect_name, psoc, cover_category) |>
  # calculate mean percent cover
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join with metadata by year_pool
  left_join(metadata, by = "year_pool")


grass_cover <- gl_implicitNA_alyssa |>
  filter(cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))

gl_transects <- gl_implicitNA_alyssa |>
  filter(
    transect_name %in% c("GL-02", "GL-07", "GL-09", "GL-10",
                         "GL-11", "GL-12", "GL-14", "GL-15"),
    cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER")
  )



```

### FINAL all pc faceted / transect
```{r}
#| label: alyssa vis-1 faceted native non native pc over time 

gl_transects |>
  group_by(year, transect_name, cover_category) |>
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  ggplot(aes(x = year,
             y = total_cover,
             color = cover_category)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ transect_name, scales = "free_x") +
  labs(
    title = "Native vs Non-Native Percent Cover Over Time Across GL Transects",
    x = "Year",
    y = "Total Percent Cover",
    color = "Cover Category"
  ) +
  scale_color_manual(
    values = c("NON-NATIVE COVER" = "#C96B5C",
               "NATIVE COVER" = "#4A9782")) +
  theme_bw()
```

### non native and native species of interest over time FACETED by species
```{r}
#| label: alyssa wrangling for vis 2 

selected_species <- c(native_interest, non_native_interest)

selected_species_cover <- gl_implicitNA |>
  filter(psoc %in% selected_species) |>
  mutate(
    species_group = case_when(
      psoc %in% native_interest ~ "Native",
      psoc %in% non_native_interest ~ "Non-native"
    )
  )
```

### FINAL key spp faceted / spp 
```{r}
#| label: alyssa vis-2 faceted species of interest over time ALL (mean pc)
#| fig-width: 14
#| fig-height: 8

selected_species_cover |>
  group_by(year, psoc, species_group) |>
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  ungroup() |>
  mutate(
    psoc = str_replace_all(psoc, "_", " "),
    species_group = factor(species_group,
                           levels = c("Native", "Non-native"))
  ) |>
  arrange(species_group, psoc) |>
  mutate(psoc = factor(psoc, levels = unique(psoc))) |>
  ggplot(aes(x = year,
             y = total_cover,
             color = species_group)) +
  geom_line(aes(group = psoc),
            linewidth = 1.2) +
  gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(
      color = "gray70",
      alpha = 0.6,
      linewidth = 0.5
    )
  ) +
  scale_color_manual(
    breaks = c("Native", "Non-native"),
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#C96B5C"
    )
  ) +
  facet_wrap(~ psoc, scales = "fixed") +
  scale_x_continuous(
    breaks = 2021:2025,
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  labs(
    title = "Selected Species Percent Cover Over Time",
    x = "Year",
    y = "Total Percent Cover",
    color = "Species Group"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 8)
  )
```

```{r}
#| label: abby revised vis-2 faceted species of interest over time ALL
#| fig-width: 20
#| fig-height: 14

# NATIVE
selected_pc_native = selected_species_cover |>
  group_by(year, psoc, species_group) |>
  filter(species_group == "Native") |> 
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  ungroup() |>
  mutate(psoc = str_replace_all(psoc, "_", " ")) |> 
  arrange(desc(total_cover)) |>
  mutate(psoc = factor(psoc, levels = unique(psoc))) |>
  ggplot(aes(x = year,
             y = total_cover,
             color = species_group)) +
  geom_line(aes(group = psoc),
            linewidth = 1.2) +
  gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(
      color = "gray70",
      alpha = 0.6,
      linewidth = 0.5
    )
  ) +
  scale_color_manual(
    breaks = c("Native"),
    values = c(
      "Native" = "#4A9782"
    )
  ) +
  facet_wrap(~ psoc, scales = "free_x", nrow = 1) +
  scale_x_continuous(
    breaks = 2021:2025,
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  labs(
    title = "Selected Native Species Percent Cover Over Time",
    x = "Year",
    y = "Total Percent Cover",
    color = "Species Group"
  ) +
  theme_bw() + 
  theme(axis.text.x = element_text(size = 8)) +
  theme(legend.position = "none")

# NON-NATIVE
selected_pc_nonnative = selected_species_cover |>
  group_by(year, psoc, species_group) |>
  filter(species_group == "Non-native") |> 
  summarize(total_cover = sum(mean_pc, na.rm = TRUE)) |>
  ungroup() |>
  mutate(psoc = str_replace_all(psoc, "_", " ")) |> 
  arrange(desc(total_cover)) |>
  mutate(psoc = factor(psoc, levels = unique(psoc))) |>
  ggplot(aes(x = year,
             y = total_cover,
             color = species_group)) +
  geom_line(aes(group = psoc),
            linewidth = 1.2) +
  gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(
      color = "gray70",
      alpha = 0.6,
      linewidth = 0.5
    )
  ) +
  scale_color_manual(
    breaks = c("Non-native"),
    values = c(
      "Non-native" = "#C96B5C"
    )
  ) +
  facet_wrap(~ psoc, scales = "free_x", nrow = 1) +
  scale_x_continuous(
    breaks = 2021:2025,
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  labs(
    title = "Selected Non-native Species Percent Cover Over Time",
    x = "Year",
    y = "Total Percent Cover",
    color = "Species Group"
  ) +
  theme_bw() + 
  theme(axis.text.x = element_text(size = 8)) + 
  theme(legend.position = "none")

selected_pc_native/selected_pc_nonnative
```

### wrangling for below 
```{r}
#| label: alyssa vis-3 species richness native non native wrangling


# create new object from gl_explicit0
# create new object from gl_implicitNA_alyssa
alyssa_species_richness <- gl_implicitNA_alyssa |> 
  # group by year, transect, and cover category
  group_by(year, transect_name, cover_category) |> 
  # filter data frame to include only native and non-native species
  # also filter percent cover to identify species presence
  filter(cover_category %in% c("NATIVE COVER",
                               "NON-NATIVE COVER"),
         mean_pc > 0) |> 
  # species richness calculation
  summarize(spp_rich = n_distinct(psoc)) |> 
  # ungroup data frame
  ungroup()
```

### FINAL all spp richness faceted / transect

```{r}
#| label: alyssa vis-3 species richness native non native boxplot + wrangling

# combine native and non-native richness data
all_species_richness <- bind_rows(
  # add native label
  native_species_rich |>
    mutate(type = "Native"),
  # add non-native label
  invasive_species_rich |>
    mutate(type = "Non-native")
)


# create combined bar plot
species_richness_bar_plot <- all_species_richness |>
  # create ggplot
  ggplot(aes(x = year,
             y = spp_rich,
             fill = type)) +
  # create side-by-side bars
  geom_col() +
  # facet by transect
  facet_wrap(~ transect_name) +
  # customize colors
  scale_fill_manual(values = c(
    "Native" = "#4A9782",
    "Non-native" = "#D16D5B"
  )) +
  # add labels
  labs(
    title = "Native vs Non-native Species Richness Over Time",
    x = "Year",
    y = "Species Richness",
    fill = "Species Type"
  ) +
  # black and white theme
  theme_bw()
# display plot
species_richness_bar_plot

```

```{r}
#| label: alyssa vis-3 native species richness barplot

# native species richness over time
native_species_rich_plot <- native_species_rich |>
  # create ggplot
  ggplot(aes(x = year,
             y = spp_rich,
             group = transect_name)) +
  # create bars
  geom_col(fill = "#4A9782") +
  # facet by transect
  facet_wrap(~ transect_name) +
  # add labels
  labs(
    title = "Native Species Richness Over Time",
    x = "Year",
    y = "Native Species Richness"
  ) +
  # black and white theme
  theme_bw()
```

```{r}
#| label: alyssa vis-3 species richness invasive barplot

# invasive/non-native species richness over time
invasive_species_rich_plot <- invasive_species_rich |>
  # create ggplot
  ggplot(aes(x = year,
             y = spp_rich,
             group = transect_name)) +
  # create bars
  geom_col(fill = "#D16D5B") +
  # facet by transect
  facet_wrap(~ transect_name) +
  # add labels
  labs(
    title = "Non-native Species Richness Over Time",
    x = "Year",
    y = "Non-native Species Richness"
  ) +
  # black and white theme
  theme_bw()
```

```{r fig.height=8}
native_species_rich_plot / invasive_species_rich_plot
```


### d plot all 
```{r}
#| label: alyssa vis-4 density plot ALL native non native richness

# combine native and non-native richness data
all_species_richness <- bind_rows(
  # add native label
  native_species_rich |>
    mutate(type = "Native"),
  # add non-native label
  invasive_species_rich |>
    mutate(type = "Non-native")
)
# create combined density plot
species_richness_density_plot <- all_species_richness |>
  # create ggplot
  ggplot(aes(x = spp_rich,
             fill = type,
             color = type)) +
  # create overlapping density plots
  geom_density(alpha = 0.4,
               linewidth = 1) +
  # facet by transect
  facet_wrap(~ transect_name) +
  # customize colors
  scale_fill_manual(values = c(
    "Native" = "#4A9782",
    "Non-native" = "#D16D5B"
  )) +
  scale_color_manual(values = c(
    "Native" = "#4A9782",
    "Non-native" = "#D16D5B"
  )) +
  # add labels
  labs(
    title = "Native vs Non-native Species Richness Distributions",
    x = "Species Richness",
    y = "Density",
    fill = "Species Type",
    color = "Species Type"
  ) +
  # black and white theme
  theme_bw()
# display plot
species_richness_density_plot
```

### FINAL REVISION - updated bar plot for all species w event
```{r}
#| fig-width: 7
#| fig-height: 5


# combine native and non-native richness data
all_species_richness <- bind_rows(
  native_species_rich |>
    mutate(type = "Native"),
  invasive_species_rich |>
    mutate(type = "Non-native")
)
# summarize species richness by year and type
species_richness_summary <- all_species_richness |>
  group_by(year, type) |>
  summarize(
    total_richness = sum(spp_rich, na.rm = TRUE),
    .groups = "drop"
  )
# create stacked bar plot over time
species_richness_bar_plot <- species_richness_summary |>
  ggplot(aes(x = year,
             y = total_richness,
             fill = type)) +
  # stacked bars
  geom_col() +
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray50",
    linewidth = 1
  ) +
  # label prescribed fire event
  annotate(
    "text",
    x = 2023.5,
    y = 184,
    label = "Prescribed Fire\n(Fall 2023)",
    angle = 90,
    vjust = -0.5,
    size = 3,
    color = "gray40"
  ) +
  scale_fill_manual(values = c(
    "Native" = "#4A9782",
    "Non-native" = "#D16D5B"
  )) +
  labs(
    title = "Native vs Non-native Species Richness Over Time",
    x = "Year",
    y = "Species Richness",
    fill = "Species Type"
  ) +
  theme_bw()

# display plot
species_richness_bar_plot
```


### d plot by selected species richness
```{r}
#| label: wrangling for alyssa vis-5 density plot focus native non native richness

# create species richness object for selected species
selected_species_richness <- selected_species_cover |>
  # filter percent cover to identify species presence
  filter(mean_pc > 0) |>
  # group by year, transect, and native/non-native group
  group_by(year, transect_name, species_group) |>
  # count number of unique species present
  summarize(spp_rich = n_distinct(psoc),
            .groups = "drop")
```

### FINAL REVISION - updated bar plot for selected species
```{r}
#| label: updated bar plot for selected species
#| fig-width: 7
#| fig-height: 5


# create species richness object for selected species
selected_species_richness <- selected_species_cover |>
  # filter percent cover to identify species presence
  filter(mean_pc > 0) |>
  # group by year, transect, and species group
  group_by(year, transect_name, species_group) |>
  # count unique species present
  summarize(
    spp_rich = n_distinct(psoc),
    .groups = "drop"
  )
# summarize richness across years
selected_species_richness_summary <- selected_species_richness |>
  group_by(year, species_group) |>
  summarize(
    total_richness = sum(spp_rich, na.rm = TRUE),
    .groups = "drop"
  )

# create stacked bar plot
selected_species_richness_bar_plot <- selected_species_richness_summary |>
  ggplot(aes(x = year,
             y = total_richness,
             fill = species_group)) +
  # create stacked bars
  geom_col() +
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray50",
    linewidth = 1
  ) +
  # label prescribed fire event
  annotate(
    "text",
    x = 2023.5,
    y = 68,
    label = "Prescribed Fire\n(Fall 2023)",
    angle = 90,
    vjust = -0.5,
    size = 3,
    color = "gray40"
  ) +
  # customize colors
  scale_fill_manual(values = c(
    "Native" = "#4A9782",
    "Non-native" = "#D16D5B"
  )) +
  # add labels
  labs(
    title = "Selected Species Richness Over Time",
    x = "Year",
    y = "Species Richness",
    fill = "Species Group"
  ) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 80)) +
  # black and white theme
  theme_bw()

# display plot
selected_species_richness_bar_plot
```

### ?final? - 4 key spp + fire event 

```{r}
#| label: key four spp wrangling

# choose four focal species
four_species <- c("Deinandra_fasciculata",
                  "Stipa_pulchra",
                  "Festuca_perennis",
                  "Polypogon_monspeliensis")

# create object with only the four focal species
four_species_cover <- selected_species_cover |>
  filter(psoc %in% four_species) |>
  group_by(year, psoc, species_group) |>
  summarize(total_cover = sum(mean_pc, na.rm = TRUE),
            .groups = "drop")
```

```{r}
#| label: stipa-jitter-plot

# Stipa pulchra jitter plot with mean points
stipa_plot <- selected_species_cover |>
  
  # keep only Stipa pulchra
  filter(psoc == "Stipa_pulchra") |>
  
  # create plot
  ggplot(aes(x = year,
             y = mean_pc)) +
  
  # blue shaded region shows wet 2024 water year
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
  
  # individual transect observations
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
  
  # yearly mean point
stat_summary(
  aes(color = "Native"),
  fill = "#4A9782", 
  geom = "point",
  fun = "mean",
  shape = 21,
  size = 5,
  stroke = 1.2
) +
  scale_color_manual(
  name = "Species Type",
  values = c(
    "Native" = "#4A9782",
    "Non-native" = "#9B5DE5"
  )
) +
  
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
  
  # label prescribed fire event
  annotate(
    "text",
    x = 2023.55,
    y = 50,
    label = "Prescribed fire\n(Fall 2023)",
    hjust = 0,
    vjust = 1,
    color = "gray35",
    size = 3.5
  ) +
  
  # show every year on x-axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
  
  # relabel axes and add title
  labs(
    title = "Stipa Pulchra",
    y = "Mean Percent Cover",
    x = NULL
  ) +
  
  # custom theme
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold",
                              size = 16),
    plot.subtitle = element_text(color = "gray35"),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

# display plot
stipa_plot
```

```{r}
#| label: deinandra-jitter-plot

#| label: deinandra-jitter-plot

# Deinandra fasciculata jitter plot with mean points
deinandra_plot <- selected_species_cover |>
  
  # keep only Deinandra fasciculata
  filter(psoc == "Deinandra_fasciculata") |>
  
  # create plot
  ggplot(aes(x = year,
             y = mean_pc)) +
  
  # blue shaded region shows wet 2024 water year
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
  
  # individual transect observations
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
  
  # yearly mean point
  stat_summary(
    aes(color = "Native"),
    fill = "#4A9782",
    geom = "point",
    fun = "mean",
    shape = 21,
    size = 5,
    stroke = 1.2
  ) +
  
  scale_color_manual(
    name = "Species Type",
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#9B5DE5"
    )
  ) +
  
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
  
  # show every year on x-axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
  
  # relabel axes and add title
  labs(
    title = "Deinandra Fasciculata",
    y = NULL,
    x = NULL
  ) +
  
  # custom theme
  theme_bw() +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

# display plot
deinandra_plot
```

```{r}
#| label: festuca-jitter-plot


# Festuca perennis jitter plot with mean points
festuca_plot <- selected_species_cover |>
  
  # keep only Festuca perennis
  filter(psoc == "Festuca_perennis") |>
  
  # create plot
  ggplot(aes(x = year,
             y = mean_pc)) +
  
  # blue shaded region shows wet 2024 water year
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
  
  # individual transect observations
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
  
  # yearly mean point
  stat_summary(
    aes(color = "Non-native"),
    fill = "#9B5DE5",
    geom = "point",
    fun = "mean",
    shape = 21,
    size = 5,
    stroke = 1.2
  ) +
  
  scale_color_manual(
    name = "Species Type",
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#9B5DE5"
    )
  ) +
  
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
  
  # show every year on x-axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
  
  # relabel axes and add title
  labs(
    title = "Festuca Perennis",
    y = "Mean Percent Cover",
    x = "Year"
  ) +
  
  # custom theme
  theme_bw() +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

# display plot
festuca_plot
```

```{r}
#| label: polypogon-jitter-plot

# Polypogon monspeliensis jitter plot with mean points
polypogon_plot <- selected_species_cover |>
  
  # keep only Polypogon monspeliensis
  filter(psoc == "Polypogon_monspeliensis") |>
  
  # create plot
  ggplot(aes(x = year,
             y = mean_pc)) +
  
  # blue shaded region shows wet 2024 water year
  annotate(
    "rect",
    xmin = 2023.75,
    xmax = 2024.75,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.18
  ) +
  
  # individual transect observations
  geom_jitter(
    width = 0.10,
    height = 0,
    shape = 21,
    size = 2.5,
    alpha = 0.55,
    color = "gray55",
    fill = "gray80"
  ) +
  
  # yearly mean point
  stat_summary(
    aes(color = "Non-native"),
    fill = "#9B5DE5",
    geom = "point",
    fun = "mean",
    shape = 21,
    size = 5,
    stroke = 1.2
  ) +
  
  scale_color_manual(
    name = "Species Type",
    values = c(
      "Native" = "#4A9782",
      "Non-native" = "#9B5DE5"
    )
  ) +
  
  # prescribed fire occurred after summer 2023 sampling
  geom_vline(
    xintercept = 2023.5,
    linetype = "dashed",
    color = "gray45",
    linewidth = 0.8
  ) +
  
  # show every year on x-axis
  scale_x_continuous(
    breaks = 2021:2025
  ) +
  
  # relabel axes and add title
  labs(
    title = "Polypogon Monspeliensis",
    y = NULL,
    x = "Year"
  ) +
  
  # custom theme
  theme_bw() +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

# display plot
polypogon_plot
```

### FINAL plot 4 key spp with events 
```{r}
#| label: focal-species-combined
#| fig-width: 10
#| fig-height: 8

combined_species_plot <-
  ((stipa_plot + deinandra_plot) /
   (festuca_plot + polypogon_plot)) +
  
  plot_layout(guides = "collect") +
  
  plot_annotation(
    title = "Percent Cover of Four Keys Species Through Time",
    subtitle = "Blue shading indicates the 2024 water year (143% of normal precipitation).",
    theme = theme(
      plot.title = element_text(
        size = 20,
        face = "bold",
        hjust = 0.5
      ),
      plot.subtitle = element_text(
        size = 12,
        color = "gray35",
        hjust = 0.5
      )
    )
  ) &
  
  theme(
    legend.position = "bottom"
  )

# display plot
combined_species_plot
```

